Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-07-08 inference.tex

The rely-guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers.  The idea is to infer the logical effect that each
statement has on the transaction-local database state $\stl$ (i.e.,
how it transforms $\stl$), and compose multiple such effects together
to describe the effect of the transaction as a whole.  Importantly,
this approach generalizes to loops, where the effect of a loop can be
inferred as a well-defined function of the effect of its body, thanks
to certain pleasant properties enjoyed by the database programs
modeled by $\txnimp$.  Interpreting database semantics as functional
transformations on sets (described in terms of their logical effects)
enables an inference mechanism that can leverage off-the-shelf SMT
solvers for automated verification.

@@

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
the supplementary material.}:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation as discussed below.
% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule so updated is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

>>>>

The rely-guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers that alleviates this burden.  The idea is to infer the
logical effect that each statement has on the transaction-local
database state $\stl$ (i.e., how it transforms $\stl$), and compose
multiple such effects together to describe the effect of the
transaction as a whole.  Importantly, this approach generalizes to
loops, where the effect of a loop can be inferred as a well-defined
function of the effect of its body, thanks to certain pleasant
properties enjoyed by the database programs modeled by $\txnimp$.
Interpreting database semantics as functional transformations on sets
(described in terms of their logical effects) enables an inference
mechanism that can leverage off-the-shelf SMT solvers for automated
verification.

@@

The correspondence between the inference rules and the RG judgment of
\S~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
the supplementary material.}:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl \subseteq \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation as discussed below.
% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule so updated is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

### 2017-07-08 isolation.tex

A distinctive characteristic of our development is that it is
parameterized on a weak isolation specification $\I$. $\I$ can be
instantiated with the declarative characterization of an isolation
guarantee or a concurrency control mechanism, regardless of the actual
implementation used to realize it. This allows us to model a range of
isolation properties that are relevant to the theory and practice of
transaction processing systems without appealing to specific
implementation artifacts like locks, versions, logs, speculation, etc.
A few well-known properties are discussed below:

@@

\textbf{Unique Ids}. As the \C{new\_order} example
(Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
of ordered identifiers requires stronger isolation levels than the
ones that are default on most databases (e.g., Read
Committed). Alternatively, globally unique sequence numbers,
regardless of the isolation level, can be requested from a relational
database via SQL's \C{UNIQUE} and \C{AUTO\_INCREMENT} keywords. Our
development crucially relies on the uniqueness of record
identifiers\footnote{The importance of unique ids is recognized in
  real-world implementations.  For example, MySQL's InnoDB engine
  automatically adds a 6-byte unique identifier if none exists for a
  record.}, which are checked locally for uniqueness by the
\rulelabel{E-Insert} rule.  The global uniqueness of locally unique
identifiers can be captured as an isolation property thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ ensures that if the id of a record is globally unique when
it is added to a transaction's $\stl$, it remains globally unique
until the transaction commits. This would be achieved within our
semantic framework by prohibiting the interference from a concurrent
transaction that adds the same id. The axiom thus simulates a global
counter protected by an exclusive lock without explicitly appealing to
an implementation artifact.

@@

\textbf{Read Committed (RC) and Monotonic Atomic View (MAV)} RC
isolation allows a transaction to witness writes of committed
transactions at any point during the transaction's execution.
Although it offers only weak isolation guarantees, it nonetheless
prevents witnessing \emph{dirty writes} (i.e., writes performed by
uncommitted transactions).  Monotonic Atomic View
(MAV)~\cite{bailishat} is an extension to RC that guarantees the
continuous visibility of a committed transaction's writes once they
become visible in the current transaction. That is, a MAV transaction
does not witness \emph{disappearing writes}, which can happen on a
weakly consistent machine. Due to the SC nature of our abstract
machine (there is always a single global database state $\stg$; not a
vector of states indexed by vector clocks), and our choice to never
violate atomicity of a transaction's writes, both RC and MAV are
already guaranteed by our semantics.  Thus, defining $\I_e$ and $\I_c$
to \emph{true} ensures RC and MAV behavior under our semantics.

\textbf{Repeatable Read (RR)} By definition, multiple reads to a
transactional variable in a Repeatable Read transaction are required
to return the same value.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenonmenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction, but not prohibiting insertion of new
records. The ANSI SQL RR specification can be axiomatized in our
framework, but it requires a minor extension to our operational
semantics to track a transaction's reads. In particular, the records
returned by \C{SELECT} should be added to the local database $\stl$,
but without changing their transaction identifiers ($\txnf$ fields),
and flush ($\gg$) should only flush the records that bear the current
transaction's identifier. With this extension, ANSI SQL RR can be
axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction (since the current transaction's records are not yet
present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.

>>>>

A distinctive characteristic of our development is that it is
parameterized on a weak isolation specification $\I$ that can be
instantiated with the declarative characterization of an isolation
guarantee or a concurrency control mechanism, regardless of the actual
implementation used to realize it. This allows us to model a range of
isolation properties that are relevant to the theory and practice of
transaction processing systems without appealing to specific
implementation artifacts like locks, versions, logs, speculation, etc.
A few well-known properties are discussed below:

@@

\textbf{Unique Ids}. As the \C{new\_order} example
(\S\ref{sec:motivation}) demonstrates, enforcing global uniqueness
of ordered identifiers requires stronger isolation levels than the
ones that are default on most databases (e.g., Read
Committed). Alternatively, globally unique sequence numbers,
regardless of the isolation level, can be requested from a relational
database via SQL's \C{UNIQUE} and \C{AUTO\_INCREMENT} keywords. Our
development crucially relies on the uniqueness of record
identifiers\footnote{The importance of unique ids is recognized in
  real-world implementations.  For example, MySQL's InnoDB engine
  automatically adds a 6-byte unique identifier if none exists for a
  record.}, which are checked locally for uniqueness by the
\rulelabel{E-Insert} rule.  The global uniqueness of locally unique
identifiers can be captured as an isolation property thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ ensures that if the id of a record is globally unique when
it is added to a transaction's $\stl$, it remains globally unique
until the transaction commits. This would be achieved within our
semantic framework by prohibiting the interference from a concurrent
transaction that adds the same id. The axiom thus simulates a global
counter protected by an exclusive lock without explicitly appealing to
an implementation artifact.

@@

\textbf{Read Committed (RC) and Monotonic Atomic View (MAV).} RC
isolation allows a transaction to witness writes of committed
transactions at any point during the transaction's execution.
Although it offers only weak isolation guarantees, it nonetheless
prevents witnessing \emph{dirty writes} (i.e., writes performed by
uncommitted transactions).  Monotonic Atomic View
(MAV)~\cite{bailishat} is an extension to RC that guarantees the
continuous visibility of a committed transaction's writes once they
become visible in the current transaction. That is, a MAV transaction
does not witness \emph{disappearing writes}, which can happen on a
weakly consistent machine. Due to the SC nature of our abstract
machine (there is always a single global database state $\stg$; not a
vector of states indexed by vector clocks), and our choice to never
violate atomicity of a transaction's writes, both RC and MAV are
already guaranteed by our semantics.  Thus, defining $\I_e$ and $\I_c$
to \emph{true} ensures RC and MAV behavior under our semantics.

\textbf{Repeatable Read (RR)} By definition, multiple reads to a
transactional variable in a Repeatable Read transaction are required
to return the same value.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenonmenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction, but which does not prohibit insertion of new
records. The ANSI SQL RR specification can be axiomatized in our
framework, but it requires a minor extension to our operational
semantics to track a transaction's reads. In particular, the records
returned by \C{SELECT} should be added to the local database $\stl$,
but without changing their transaction identifiers ($\txnf$ fields),
and flush ($\gg$) should only flush the records that bear the current
transaction's identifier. With this extension, ANSI SQL RR can be
axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction (since the current transaction's records are not yet
present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.

### 2017-07-08 reasoning.tex

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level consistency conditions $I$,
on an implementation that satisfies the isolation specifications
($\I$) of its transactions\footnote{Note the difference between $I$
  and $\I$. The former constitute \emph{proof} \emph{obligations} for
  the programmer, whereas the latter describes a transaction's
  \emph{assumptions} about the operational characteristics of the
  underlying system.}.  Our proof system is essentially an adaptation
of rely-guarantee reasoning framework~\cite{rgjones} to the setting of
weakly isolated database transactions.  The primary challenge in the
formulation deals with how we relate a transaction's isolation
specification ($\I$) to its rely relation ($R$) that describes the
transaction's environment, so that interference is considered only
insofar as allowed by the isolation level.  Another characteristic of
the transaction setting that affects the structure of the proof system
is atomicity; a transaction's writes are not visible to its peers
until it commits.  In the context of rely-guarantee, this means that
the transaction's guarantee ($G$) should capture the aggregate effect
of a transaction, and not its individual writes.  While shared memory
\C{atomic} blocks also have the same characteristic, the fact that
transactions are weakly isolated introduces complexity.  Unlike an
\C{atomic} block, the effect of a transaction is \emph{not} a
sequential composition of the effects of its statements because each
statement witnesses a potentially different version of the state.

@@

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_e$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}

While $R$ describes changes in the global database state, $\I_e$
constrain the global effects that will visible to the transaction.
Recall, however that $\I_e$ uses the global state in effect at the
point the transaction last took a step ($\stg_1$) to determine whether
the transaction can proceed from the current global state ($\stg$).
Moreover, the global rely relation $R$ defines the permissible
interferences from $\Delta$ to a future global state ($\Delta'$).  The
local rely relation $\R_e$ determines allowed interferences as
specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
$\stg_1$ defines this reley relation to be an over-approximation on
the interferences in the presence of which the transaction can
proceed, simplifying verification burden.

@@

Salient rules of transaction-local RG judgment are shown in
Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},
\rulelabel{RG-Select}, \rulelabel{RG-Delete}, and
\rulelabel{RG-Insert}) reflect the structure of the corresponding
reduction rule from Fig.~\ref{fig:txnimp}.  The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $\psi$ that needs to be stable w.r.t $\R$. $\psi$ must be
implied by $P$, the pre-condition of \C{FOREACH}, when no elements
have been iterated, i.e, when $y=\emptyset$. The body of the loop can
assume the loop invariant, and the fact that $z$ is an element from
the set $x$ being iterated, to prove its post-condition $Q_c$. The
operational semantics ensures that $z$ is added to $y$ at the end of
the iteration, hence $Q_c$ must imply $[y\cup\{z\}/y]\psi$. When the
loop has finished execution, $y$, the set of iterated items, is the
entire set $x$. Thus $[x/y]\psi$ is true at the end of the loop, from
which the post-condition $Q$ must follow. As with the other rules, $Q$
needs to be stable. The rules for conditionals, sequencing etc., are
standard, and hence elided.

>>>>

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level consistency conditions $I$,
on an implementation that satisfies the isolation specifications
($\I$) of its transactions\footnote{Note the difference between $I$
  and $\I$. The former constitute \emph{proof} \emph{obligations} for
  the programmer, whereas the latter describes a transaction's
  \emph{assumptions} about the operational characteristics of the
  underlying system.}.  Our proof system is essentially an adaptation
of a rely-guarantee reasoning framework~\cite{rgjones} to the setting of
weakly isolated database transactions.  The primary challenge in the
formulation deals with how we relate a transaction's isolation
specification ($\I$) to its rely relation ($R$) that describes the
transaction's environment, so that interference is considered only
insofar as allowed by the isolation level.  Another characteristic of
the transaction setting that affects the structure of the proof system
is atomicity; a transaction's writes are not visible to its peers
until it commits.  In the context of rely-guarantee, this means that
the transaction's guarantee ($G$) should capture the aggregate effect
of a transaction, and not its individual writes.  While shared memory
\C{atomic} blocks also have the same characteristic, the fact that
transactions are weakly isolated introduces surprising complexity.  Unlike an
\C{atomic} block, the effect of a transaction is \emph{not} a
sequential composition of the effects of its statements because each
statement can witness a potentially different version of the state.

@@

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It depends on a
transaction-local judgment to verify the body ($c$) of a transaction with
id $i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_e$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
\noindent While $R$ describes changes in the global database state, $\I_e$
constrains the global effects visible to the transaction.
Recall however that $\I_e$ uses the global state in effect at the
point the transaction last took a step ($\stg_1$) to determine whether
the transaction can proceed from the current global state ($\stg$).
Moreover, the global rely relation $R$ defines the permissible
interferences from $\Delta$ to a future global state ($\Delta'$).  The
local rely relation $\R_e$ determines allowed interferences as
specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
$\stg_1$ defines this reley relation to be an over-approximation on
the interferences in the presence of which the transaction can
proceed, simplifying verification burden.

@@

Salient rules of transaction-local RG judgments are shown in
Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},
\rulelabel{RG-Select}, \rulelabel{RG-Delete}, and
\rulelabel{RG-Insert}) reflect the structure of the corresponding
reduction rule from Fig.~\ref{fig:txnimp}.  The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $\psi$ that needs to be stable w.r.t $\R$. $\psi$ must be
implied by $P$, the pre-condition of \C{FOREACH}, when no elements
have been iterated, i.e, when $y=\emptyset$. The body of the loop can
assume the loop invariant, and the fact that $z$ is an element from
the set $x$ being iterated, to prove its post-condition $Q_c$. The
operational semantics ensures that $z$ is added to $y$ at the end of
the iteration, hence $Q_c$ must imply $[y\cup\{z\}/y]\psi$. When the
loop has finished execution, $y$, the set of iterated items, is the
entire set $x$. Thus $[x/y]\psi$ is true at the end of the loop, from
which the post-condition $Q$ must follow. As with the other rules, $Q$
needs to be stable. The rules for conditionals, sequencing etc., are
standard, and hence elided.
